---
title: "Integrative sparse Partial Least Squares Discriminant Analysis multiomics modeling"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
author: Michael Chimenti
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction 

**This is an attempt to use the RNA-seq and Proteome data to create a predictive model of COVID severity.**  We will use the sPLS method as before, but with an additional "DA" (discriminant analysis) term. We are interested in evaluating correlations and relationships between the RNA-seq and proteomic data in our 4 covid cohorts (COVID/ICU, COVID/Inpatient, COVID/Outpatient, No-COVID,Inpatient).  We are looking at the two-way integration before attempting the 3-way modeling incorporating the exome data. PLS was designed with a canonical (exploratory) approach and a regression (explanatory) approach in mind. Partial Least Squares – Discriminant Analysis (PLS-DA) was hence developed to allow the powerful PLS algorithm to be used for classification [1, 2]. It performs very similarly to PLS, just that the response vector y contains categorical vectors rather than continuous vectors. PLS-DA has the same advantages that PLS does, such that it operates efficiently over large dataframes and is not negatively influenced by collinearity.

# Methods 

We are following the general pattern in this vignette (http://mixomics.org/methods/spls-da/) and this case study (http://mixomics.org/case-studies/splsda-srbct-case-study/).    

# Analysis 

## Imports

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(here)
library(mixOmics)

here::i_am("analysis/mixomics_sPLS_DA_predictive.Rmd")
```

## Load data 

```{r}

rna <- readr::read_csv(here("data","munged_data","rnaseq_samples_in_rows_normalized.csv"))
prot <- readr::read_csv(here("data","munged_data","proteomics_PLAS_ONLY_samples_in_rows_norm_NAMEFIX.csv"))
meta <- readr::read_csv(here("data","munged_data","metadata_masterkey.csv"))
```

## Wrangle metadata

Make the sample names match.  Not all samples specified in the metadata are found in the datafiles.  

```{r}
new_name <- rna$name %>% str_split(., pattern="-", simplify = TRUE) %>% paste0("RL_SARS_",.)
new_name <- new_name[1:80]
rna$sample_ID <- new_name 

prot <- prot %>% rename(sample_ID=Assay)

### There are 76 of 168 total proteomic datasets (we are excluding saliva datasets from consideration which is the reason for the mismatch)
sum(prot$sample_ID %in% meta$sample_ID)

###  There are 70 of 100 RNAseq files specified in meta (I don't know where the missing 30 files are)
sum(rna$sample_ID %in% meta$sample_ID)

### There are 30 samples listed in metadata NOT found among the RNA data files 
sum(meta$sample_ID[meta$sample_type=="RNA"] %in% rna$sample_ID)

## These missing samples are: 
missing <- meta$sample_ID[meta$sample_type=="RNA"] %in% rna$sample_ID
no_RNA_data <- meta[meta$sample_type=="RNA",] %>% filter(!missing)
no_RNA_data$sample_ID

### Each data table needs a "master key" to determine sample of origin, without regards to data type 
master_key_dict <- meta %>% dplyr::select(c(sample_ID, master_key))
prot2 <- left_join(prot, master_key_dict, by="sample_ID") %>% drop_na()
rna2 <- left_join(rna, master_key_dict, by="sample_ID")

### We are missing some data...need to confirm with Ryan lab 
intersect(prot2$master_key, rna2$master_key)
setdiff(prot2$master_key, rna2$master_key)

## Note: 
## RNA files for 5.7 through 7.1 are in 'Auto_user_GSS5PR-0221-28-first_RNA_run_198.mean_hk_scaled_log2_count.xls'
## Need to edit RNA munging to include these and find location on samples 25-29, 30-34,... etc... 
```


### Convert data tables to named matrices 

```{r}
match <- intersect(prot2$master_key, rna2$master_key)
prot3 <- prot2[prot2$master_key %in% match,] %>% dplyr::select(-c(sample_ID)) %>% as.data.frame()
rownames(prot3) <- prot3$master_key
prot3 <- prot3 %>% dplyr::select(-master_key)

rna3 <- rna2[rna2$master_key %in% match,] %>% dplyr::select(-c(name, sample_ID)) %>% as.data.frame()
rownames(rna3) <- rna3$master_key
rna3 <- rna3 %>% dplyr::select(-master_key)

## check dimensions match 
dim(rna3)
dim(prot3)

## The metadata categories are the same whether RNA or Plasma 
meta_3 <- meta_3 <- meta[meta$master_key %in% match & meta$sample_type=="RNA",]

## Sanity check 
rownames(rna3) == meta_3$master_key
rownames(prot3) == meta_3$master_key
```

### Initial sPLS-DA model 

For starters, we'll include the COVID NEG samples just to see how the model looks: 

```{r}
spls.da.rna.covid <- splsda(X=rna3, Y=meta_3$cohort_status, ncomp=10)
spls.da.pro.covid <- splsda(X=prot3, Y=meta_3$cohort_status, ncomp=10)

```

```{r}
# plot the samples projected onto the first two components of the PLS-DA subspace from RNA-seq data
plotIndiv(spls.da.rna.covid , comp = 1:2, 
          group = meta_3$cohort_status, ind.names = FALSE,  # colour points by class
          ellipse = TRUE, # include 95% confidence ellipse for each class
          legend = TRUE, title = 'RNA-seq PLSDA with confidence ellipses')


plotIndiv(spls.da.pro.covid , comp = 1:2, 
          group = meta_3$cohort_status, ind.names = FALSE,  # colour points by class
          ellipse = TRUE, # include 95% confidence ellipse for each class
          legend = TRUE, title = 'Proteome PLSDA with confidence ellipses')

# use the max.dist measure to form decision boundaries between classes based on PLS-DA data
background = background.predict(spls.da.rna.covid, comp.predicted=2, dist = "max.dist")

# plot the samples projected onto the first two components of the PLS-DA subspace
plotIndiv(spls.da.rna.covid, comp = 1:2,
          group = meta_3$cohort_status, ind.names = FALSE, # colour points by class
          background = background, # include prediction background for each class
          legend = TRUE, title = " (b) PLSDA with prediction background")
```

### sPLS-DA without the COVID NEG samples 

```{r}
mask <- meta_3$covid_status != "NEG"
rna4 <- rna3[mask,]
prot4 <- prot3[mask,]
meta_4 <- meta_3[mask,]

## check dimensions match 
dim(rna4)
dim(prot4)

## Sanity check 
rownames(rna4) == meta_4$master_key
rownames(prot4) == meta_4$master_key
```

```{r}

spls.da.rna.covid <- splsda(X=rna4, Y=meta_4$cohort_status, ncomp=10)
spls.da.pro.covid <- splsda(X=prot4, Y=meta_4$cohort_status, ncomp=10)

# plot the samples projected onto the first two components of the PLS-DA subspace from RNA-seq data
plotIndiv(spls.da.rna.covid , comp = 1:2, 
          group = meta_4$cohort_status, ind.names = FALSE,  # colour points by class
          ellipse = TRUE, # include 95% confidence ellipse for each class
          legend = TRUE, title = 'RNA-seq PLSDA with confidence ellipses')


plotIndiv(spls.da.pro.covid , comp = 1:2, 
          group = meta_4$cohort_status, ind.names = FALSE,  # colour points by class
          ellipse = TRUE, # include 95% confidence ellipse for each class
          legend = TRUE, title = 'Proteome PLSDA with confidence ellipses')

# use the max.dist measure to form decision boundaries between classes based on PLS-DA data
background_rna = background.predict(spls.da.rna.covid, comp.predicted=2, dist = "max.dist")

# plot the samples projected onto the first two components of the PLS-DA subspace
plotIndiv(spls.da.rna.covid, comp = 1:2,
          group = meta_4$cohort_status, ind.names = TRUE, # colour points by class
          background = background_rna, # include prediction background for each class
          legend = TRUE, title = " RNA-seq PLSDA with prediction background")


background_pro = background.predict(spls.da.pro.covid, comp.predicted=2, dist = "max.dist")
plotIndiv(spls.da.pro.covid, comp = 1:2,
          group = meta_4$cohort_status, ind.names = TRUE, # colour points by class
          background = background_pro, # include prediction background for each class
          legend = TRUE, title = " Proteome PLSDA with prediction background")

```

NB: "s22_day1" is an outlier for the COVID POS Inpatient cohort in Proteomics, "s24_day1" is an outlier for RNA-seq.  

```{r}



```

